Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-06-12 cordillera-cycle.tex
Add corrections from Irina.

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving atmospheric
  forcing, all simulations produce two major glaciation events during MIS~4
  (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}). The timing of
  glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional ocean sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remains of Cordilleran ice during the
  early Holocene (10.9--9.5\,\unit{kyr}).

@@

\begin{center}
\begin{tabular}{>{\bfseries}ll}
    \tophline
    Introduction & Happy with it except for missing refs.\\
    Model setup  & Happy with it except for missing refs.\\
    Results      & Happy with it except for missing refs.\\
    Discussion   & Deglaciation parts needs improvement.\\
    Conclusions  & Can be compacted and the style improved.\\
    \bottomhline
\end{tabular}
\end{center}

@@

Independently of the choice of palaeo-temperature record used as forcing
time-series for the ice sheet model, our simulations consistently produce two
glaciation maxima during MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2
(29.5--16.9\,\unit{kyr}). These events broadly correspond in timing to the
Gladstone (MIS~4) and McConell (MIS~2) glaciation documented by geological
evidence in the northern part of the model domain, and to the Fraser (MIS~2)
glaciation in the southern part.

However, the timing of the modelled glacial maxima strongly depends on the
choice of palaeo-temperature record used, and a more precise agreement with
timing reconstructed from geological evidence is only attained in some of our
simulations. In the Puget lowland, the last glacial maximum advance of the
southern Cordilleran ice sheet margin has been constrained between by
radiocarbon dating between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}.
These dates have been confirmed by the offshore sediment record, which shows an
increase of glaciomarine sedimentation between 19.5 and
16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}.

Among simulations presented here, only those forced by the GRIP, EPICA and
Vostok records place the last glacial maximum within this range
(Fig.~\ref{fig:timeseries}, lower panel). Simulations driven by the NGRIP,
ODP~1012 and ODP~1020 records, on the opposite, produce a last glacial maximum
that pre-dates reconstructed ages by several thousand years. Concerning
simulations driven by oceanic (ODP~1012, ODP~1020) records, this early
deglaciation is caused by an early warming in the palaeo-temperature records
(Fig.~\ref{fig:timeseries}, upper panel). However, it has been shown that this
early warming is a local effect, corresponding to a weakening of the California
current \aref. The California current, driving cold waters southwards
along the western coast of North America, is thought to have weakened in the
presence of the Cordilleran ice sheet, resulting in paradoxically warmer
sea-surface temperatures at the location of the ODP~1012 and ODP~1020 sites
during the last glacial maximum \aref.

@@

These results makes us believe that a palaeo-temperature proxy record located
in the direct vicinity of the former Cordilleran ice sheet would be of great
help to understand its glacial history. Because, to our knowledge, no such
record yet exists, we focus the rest of our analysis on simulations forced by
temperature records from a Greenland (GRIP) and an Antarctic (EPICA) ice cores.

@@

Although snapshots from the model output (Fig.~\ref{fig:duration}) give a
crisp picture of modelled ice dynamics at specific time frames, they often
can not be directly compared to geomorphological evidence, which is by nature
time-transgressive. In this section, we use the 6-km horizontal-resolution
model output from simulations driven by the GRIP and EPICA records to produce
a series of time-integrated plots containing information aggregated over the
whole simulation period.

During most of the glacial cycle, modelled ice cover is restricted to disjoint
ice caps centred on major mountain ranges of the North American Cordillera
(Fig.~\ref{fig:duration}). In both the GRIP and the EPICA simulation,
a~contiguous ice cover spanning from the Alaska Range to the Rocky Mountains
exists for about 29\,\unit{kyr}. This value represent cumulative ice cover over
the two glaciations during MIS~2 and MIS~4, while a dynamically consistent
Cordilleran ice sheet forms for even shorter time spans.

With exception of the oceanic front and the northern foothills of the Alaska
Range, most of the maximal extent of the ice sheet is associated with short
durations of ice cover, generally a few thousand years, distributed over one
(GRIP) or two (EPICA) glaciations. This shows that, in our simulations, the
most extensive stages of the Cordilleran ice sheet are both short-lived and
out of balance with contemporary climate states, a result which confirms
inferences from earlier geological studies \aref.

However, our simulations depict, independently of the choice of record used for
palaeo-temperature forcing, a central ice cap persisting throughout the entire
last glacial cycle on the Skeena mountains (Fig.~\ref{fig:snapshots} and
\ref{fig:duration}). In particular, this ice cap persists between the MIS~4
and MIS~2 glaciations, providing a nucleation centre for the onset of glacial
readvance towards the last glacial maximum.

During time periods when a fully-grown Cordilleran ice sheet is absent from our
simulation, the presence of the Skeena ice cap can be explained by more
widespread winter precipitation there than in other parts of the modelling
domain (Figs.~\ref{fig:atm}). Along most of the north-western North American
shore, coastal mountain ranges provide a sharp orographic obstacle to westerly
winds, capturing atmospheric moisture in the form of precipitation, and leaving
interior lowlands arid. However, near the centre of our modelling domain, the
topographic barrier formed by the Coast Mountains is less pronounced than
elsewhere, allowing westerly winds to carry moisture further inland, until it
is progressively captured by the broad and mildly elevated group of the Skeena
Mountains.

In addition, the erosional landscape of the Skeena mountain bears one of the
strongest glacial imprint (?) found within the area formerly covered by the
Cordilleran ice sheet \aref. Here we suggest that persistent ice cover
associated with basal temperatures at the ice melting point
(Fig.~\ref{fig:warmbase}) may partly explain this spectacular landscape.

@@

In our simulations, the timing of the last glacial maximum, and in turn that of
the deglaciation, depends on the palaeo-temperature time-series used to force
the model (Figs.~\ref{fig:timeseries} and ~\ref{fig:deglacseries}). The
6\,\unit{km}-resolution simulation forced by the EPICA record reaches a maximal
ice volume at 17.3\,\unit{kyr}, followed by a continuous deglaciation between
about 17 and 11\,\unit{kyr}. That driven by the GRIP record reaches maximal
volume at 19.1\,\unit{kyr}, and produce a deglaciation in three steps,
including a first stand-still between 16.6 and 14.5\,\unit{kyr}, and a late
glacial re-advance between 12.6 and 11.6\,\unit{kyr}.

Although the timing differs, the patterns of glacier retreat are relatively
consistent between the two simulations (Figs.~\ref{fig:deglac}). In both
simulations, the southern half of the modelling domain, including the Puget
Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
Columbia, are completely deglaciated at a time where there remain significant
ice cover on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
in the northern half of the domain. At this stage, a late glacial re-advance
occurs in the GRIP simulation. Both simulations then show an unzipping of the
Liard Lowland, followed by ice retreat towards the Selwyn Mountains and the
Skeena Mountains, which host the last remain of the palaeo-ice sheet.

On the Interior Plateau of British Columbia, modelled deglaciation is rapid
and northwards, resulting in the presence of a slow-flowing ice lobe on the
plateau, while the surrounding mountain ranges are already deglaciated
(Fig.~\ref{fig:profiles}). Although this supports the top-down (?)
deglaciation hypothesis for central British Columbia, modelled, southwards,
last flow directions do not agree with the local lineation patterns
(Fig.~\ref{fig:lastflow}). This could indicate that the lineations are older,
or that the modelled last glacial maximum ice sheet is too extensive to the
east, resulting in modelled deglaciation flow patterns different from those
observed in the geomorphology.

@@

Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle presented in this study consistently produce two glaciation maxima during
MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}), two periods
corresponding to documented palaeo-glaciation. This result is independent
of the choice of palaeo-temperature record used for climate forcing, and
can be seen as a first-order agreement between the model and geomorphological
evidence. However, the timing of the two glaciation peaks highly depends on
which record
is used to drive the model. The timing of the last glacial maximum is best
reproduced by the Antarctic record, and occurs too early in all simulations
that were driven by other records. The mismatch is greatest when forcing the
model with oceanic records from the Pacific Northwest, which are affected by a
weakening of the California current during the last glacial maximum.

In all simulations presented here, ice cover is limited to disjoint mountain ice
caps during most of the glacial cycle. This agrees with earlier interpretations
of the geological evidence preceding the last glacial maximum. However, our
simulations produce persistent ice cover over the Skeena mountains during
the entire glacial cycle. At time when the Cordilleran ice sheet is absent
the Skeena ice cap appears to be fed by an eastwards precipitation
intrusion through a topographic window in the Coast Mountains. The ice cap acts
as a nucleation centre at the onset of the last glacial maximum re-advance, and
may have contributed to the spectacular glacial imprint of the Skeena Mountains
erosional landscape.

Concerning the deglaciation phase, when most geological evidence is available,
none of the palaeo-temperature records used produce an ideal agreement between
model results and geological evidence. The
timing of the last glacial maximum and early deglaciation is generally best
reproduced by forcing the model with an Antarctic
record. However, the Younger Dryas re-advance is best reproduced at places
where it has been documented using temperature forcing from the GRIP record.
Nonetheless, the patterns of deglaciation are consistent between different
simulations. They show a rapid deglaciation of the southern half of the ice
sheet, followed by unzipping of the Liard Lowland, and concentric retreat of
the ice margin towards the last palaeo-ice cover in the Skeena Mountains
during the early Holocene (10.9--9.5\,\unit{kyr}).

One must keep in mind, however, that these results are only accurate for our
choices of ice-sheet model (PISM), surface mass balance model (PDD), study
object (the Cordilleran ice sheet). Most importantly, our simplistic
palaeo-climate forcing does not include precipitation corrections in response
to the presence of the ice sheet, likely leading to overestimates of glacial
extent and volume. Nevertheless, our results identify the largely understudied
Skeena Mountains as a key area to understanding glacial dynamics of the
Cordilleran ice sheet, highlighting the need for further geological
investigation in this region.

@@

% fig:warmbase
\begin{figure}
  \includegraphics{warmbase}
  \caption{Modelled duration of warm-based ice cover during the last
           120\,\unit{kyr}. Long ice cover durations combined with basal
           temperatures at the melting point may explain the strong glacial
           erosional imprint of the Skeena Mountains (SM) landscape. Hatches
           indicate areas that were covered by cold ice only.
           \note{indicate location of Skeena Mountains.}}
  \label{fig:warmbase}
\end{figure}

@@

% fig:deglac
\begin{figure}
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
           advance exceeded the last glacial maximum advanced are marked in
           green. Hatches denote the Younger Dryas re-advance, which is most
           significant in the GRIP-driven simulation.}
  \label{fig:deglac}
\end{figure}

>>>>

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving atmospheric
  forcing, all simulations produce two major glaciation events during
  marine isotope stages~4 (61.9--55.4\,\unit{kyr}) and~2
  (29.5--16.9\,\unit{kyr}). The timing of glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional ocean sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remains of Cordilleran ice during the
  early Holocene (10.9--9.5\,\unit{kyr}).

@@

\begin{center}
\begin{tabular}{>{\bfseries}ll}
    \tophline
    Introduction & Happy with it except for missing refs.\\
    Model setup  & Happy with it except for missing refs.\\
    Results      & Happy with it except for missing refs.\\
    Discussion   & Deglaciation part needs improvement.\\
    Conclusions  & Can be compacted and the style improved.\\
    \bottomhline
\end{tabular}
\end{center}

@@

Independently of the choice of palaeo-temperature record used as forcing
time-series for the ice sheet model, our simulations consistently produce two
glaciation maxima during MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2
(29.5--16.9\,\unit{kyr}). These events broadly correspond in timing to the
Gladstone (MIS~4) and McConell (MIS~2) glaciations documented by geological
evidence in the northern part of the model domain \aref, and to the Fraser
(MIS~2) glaciation in the southern part \aref.

However, the timing of the modelled glacial maxima strongly depends on the
choice of palaeo-temperature record used, and a more precise agreement with
timing reconstructed from geological evidence is only attained in some of our
simulations. In the Puget lowland, the last glacial maximum advance of the
southern Cordilleran ice sheet margin has been constrained by radiocarbon
dating between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \aref.
These dates have been confirmed by the offshore sediment record, which shows an
increase of glaciomarine sedimentation between 19.5 and
16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \aref.

Among simulations presented here, only those forced by the GRIP, EPICA and
Vostok records place the last glacial maximum within this range
(Fig.~\ref{fig:timeseries}, lower panel). Simulations driven by the NGRIP,
ODP~1012 and ODP~1020 records, on the contrary, produce a last glacial maximum
that pre-dates reconstructed ages by several thousand years. Concerning the
simulations driven by oceanic (ODP~1012, ODP~1020) records, this early
deglaciation is caused by an early warming sealed in the palaeo-temperature records
(Fig.~\ref{fig:timeseries}, upper panel). However, it has been shown that this
early warming is a local effect, corresponding to a weakening of the California
current \aref. The California current, driving cold waters southwards
along the western coast of North America, is thought to have weakened in the
presence of the Cordilleran ice sheet, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
during the last glacial maximum \aref.

@@

These results make us believe that a palaeo-temperature proxy record located
in the direct vicinity of the former Cordilleran ice sheet would be of great
help to understand its glacial history. Because, to our knowledge, no such
record yet exists, we focus the rest of our analysis on simulations forced by
temperature records from Greenland (GRIP) and Antarctic (EPICA) ice cores.

@@

Although snapshots from the model output (Fig.~\ref{fig:snapshots}) give a
crisp picture of modelled ice dynamics at specific time frames, they often
can not be directly compared to geomorphological evidence, which is by nature
time-transgressive. In this section, we use the 6-km horizontal-resolution
model output from simulations driven by the GRIP and EPICA records to produce
a series of time-integrated plots containing information aggregated over the
whole simulation period.

During most of the glacial cycle, modelled ice cover is restricted to disjoint
ice caps centred on major mountain ranges of the North American Cordillera
(Fig.~\ref{fig:duration}). In both simulations driven by the GRIP and EPICA
records, a~contiguous ice cover spanning from the Alaska Range to the Rocky
Mountains exists for about 29\,\unit{kyr} in total over MIS~2 and~4.

With the exception of the oceanic front and the northern foothills of the Alaska
Range, most of the maximal extent of the ice sheet is associated with short
durations of ice cover, generally a few thousand years, distributed over one
(GRIP) or two (EPICA) glaciations. This shows that, in our simulations, the
most extensive stages of the Cordilleran ice sheet are both short-lived and
out of balance with contemporary climate states, which confirms
inferences from earlier geological studies \aref.

However, our simulations depict, independently of the choice of record used for
palaeo-temperature forcing, a central Skeena mountain ice cap persisting
throughout the entire last glacial cycle (Fig.~\ref{fig:snapshots} and
\ref{fig:duration}). In particular, this ice cap persists between the MIS~4
and MIS~2 glaciations, providing a nucleation centre for the onset of glacial
readvance towards the last glacial maximum.

During time periods when a fully-grown Cordilleran ice sheet is absent from our
simulation, the presence of the Skeena ice cap can be explained by a more
widespread winter precipitation there than in other parts of the modelling
domain (Fig.~\ref{fig:atm}). Along most of the north-western North American
shore, coastal mountain ranges provide a sharp orographic barrier to westerly
winds, capturing atmospheric moisture in the form of precipitation, and leaving
interior lowlands arid. However, near the centre of our modelling domain, this
barrier is less pronounced than
elsewhere, allowing westerly winds to carry moisture further inland, until it
is progressively captured by the broad and mildly elevated group of the Skeena
Mountains.

In addition, the erosional landscape of the Skeena mountain bears one of the
strongest glacial imprint (?) found within the area formerly covered by the
Cordilleran ice sheet \aref. Here we suggest that persistent ice cover
associated with basal ice temperatures at the pressure-melting point
(Fig.~\ref{fig:warmbase}) may partly explain this spectacular landscape.

@@

In our simulations, the timing of the last glacial maximum, and in turn that of
the deglaciation, depends on the palaeo-temperature time-series used to force
the model (Figs.~\ref{fig:timeseries} and ~\ref{fig:deglacseries}). The
6\,\unit{km}-resolution simulation forced by the EPICA record reaches a maximal
ice volume at 17.3\,\unit{kyr}, followed by a continuous deglaciation between
about 17 and 11\,\unit{kyr}. That driven by the GRIP record reaches maximal
volume at 19.1\,\unit{kyr}, and produces a deglaciation in three steps,
including the first stand-still between 16.6 and 14.5\,\unit{kyr}, and a late
glacial re-advance between 12.6 and 11.6\,\unit{kyr}.

\todo{discuss the work that supports YD re-advance.}

Although the predicted timing of glacier retreat differs, its patterns are
relatively consistent between the two simulations (Fig.~\ref{fig:deglac}). In
both runs, the southern half of the modelling domain, including the Puget
Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
Columbia, are completely deglaciated by 12\,\unit{kyr}, whereas significant
ice cover remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
in the northern half of the domain. At this stage, a late glacial re-advance
occurs in the GRIP simulation. Both simulations then show an unzipping of the
Liard Lowland, followed by ice retreat towards the Selwyn Mountains and the
Skeena Mountains, which host the last remains of the palaeo-ice sheet.

\todo{discuss existing evidence (Martin).}

On the Interior Plateau of British Columbia, modelled deglaciation consists of
a rapid northwards retreat, leaving a slow-flowing ice lobe on the
plateau, while the surrounding mountain ranges are already deglaciated
(Fig.~\ref{fig:profiles}). Although this supports the top-down (?)
deglaciation hypothesis for central British Columbia, modelled, southwards,
last flow directions do not agree with the local lineation patterns
(Fig.~\ref{fig:lastflow}). This could indicate that the lineations are older,
or that the modelled last glacial maximum ice sheet is too extensive to the
east, resulting in modelled deglaciation flow patterns different from those
observed in the geomorphology.

@@

Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle presented in this study consistently produce two glaciation maxima during
MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}), two periods
corresponding to documented extensive palaeo-glaciations. This result is
independent of the choice of palaeo-temperature record used to approximate the
past evolution of climate forcing, and
can be seen as a first-order agreement between the model and geomorphological
evidence. However, the timing of the two glaciation peaks depends sensitively
on which record
is used to drive the model. The timing of the last glacial maximum is best
reproduced by the Antarctic record, and occurs too early in all simulations
that are driven by other records. The mismatch is greatest when using
oceanic records from the Pacific Northwest, which are affected by the
weakening of the California current during the last glacial maximum.

In all simulations presented here, ice cover is limited to disjoint mountain ice
caps during most of the glacial cycle. This agrees with earlier interpretations
of the geological evidence preceding the last glacial maximum. However, our
simulations produce persistent ice cover over the Skeena mountains during
the entire glacial cycle. At the time when a full-size Cordilleran ice sheet is
absent, the Skeena ice cap appears to be fed by the eastwards precipitation
intrusion through a topographic window in the Coast Mountains. The ice cap acts
as a nucleation centre at the onset of the last glacial maximum re-advance, and
may have contributed to the spectacular glacial imprint of the Skeena Mountains
erosional landscape.

Concerning the deglaciation phase, when most geological evidence is available,
none of the palaeo-temperature records used produces an ideal agreement between
model results and geological evidence. The timing of the last glacial maximum
and early deglaciation is generally best reproduced by forcing the model with
Antarctic records. However, the Younger Dryas re-advance is only reproduced
using temperature forcing from the GRIP record at places where it has been
documented. Nonetheless, the patterns of deglaciation are consistent between
both simulations driven by the GRIP and EPICA record. They show a rapid
deglaciation of the southern half of the ice
sheet, followed by unzipping of the Liard Lowland, and concentric retreat of
the ice margin towards the last palaeo-ice cover in the Skeena Mountains
during the early Holocene (10.9--9.5\,\unit{kyr}).

One must keep in mind, however, that these results are only accurate for our
choices of ice-sheet model (PISM), surface mass balance model (PDD) and climate
forcing. Most importantly, our simplistic
palaeo-climate forcing does not include precipitation corrections in response
to the presence of an ice cover, likely leading to overestimated glacial
extent and volume. Nevertheless, our results identify that the largely
understudied Skeena Mountains are a key area to understanding glacial dynamics
of the Cordilleran ice sheet, highlighting the need for further geological
investigation of this region.

@@

% fig:warmbase
\begin{figure}
  \includegraphics{warmbase}
  \caption{Modelled duration of warm-based ice cover during the last
           120\,\unit{kyr}. Long ice cover durations combined with basal
           temperatures at the pressure-melting point may explain the strong
           glacial erosional imprint of the Skeena Mountains (SM) landscape.
           Hatches indicate areas that were covered by cold ice only.
           \note{indicate location of Skeena Mountains.}}
  \label{fig:warmbase}
\end{figure}

@@

% fig:deglac
\begin{figure}
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
           advance exceeded the last glacial maximum advanced are marked in
           green. Hatches denote the Younger Dryas re-advance, which is more
           pronounced in the GRIP-driven simulation.}
  \label{fig:deglac}
\end{figure}
